---
title: "Plastic Degradation Term Project"
author:
- Heather MacTavish^1^*
- Douglas A. Campbell^1^*
- Mireille Savoie^1^*
- Laurel Genge^1^
- Carlie Barnhill^1^
- Max Berthold^1^

date: "`r format(Sys.Date())`"
output:
  rmdformats::readthedown:
    df_print: paged
    code_folding: hide
    self_contained: true
    keep_md: yes
    fig_caption: yes

bibliography: ../Docs/PBATDegradationCitations.bib
csl:  ../Docs/apa-with-abstract.csl
editor_options: 
  markdown: 
    wrap: 72
---
```{css, echo=FALSE}
p.caption {
  font-size: 18px;
}
```

# Affiliations {-}
^1^Mount Allison University, New Brunswick, Canada  

*corresponding author 


# Acknowledgement {-}
This R markdown file is adapted from the `Import_MCData` file created by Laurel Genge, Carlie Barnhill, Max Berthold, Mireille Savoie, and Douglas A. Campbell [@genge_import_mcdata_2024] for the Plastic Degradation project by Heather MacTavish, Andrew Forrest, and Yvanna Tchatchoua in BIOL3111.

Dr. Campbell was supported by the Canada Research Chairs.


# Overview

According to Genge et al. (2024):

>This .Rmd Rworkbook imports data in simple .csv long form exported from PSI Multicultivators based upon project specific values for variables set by the user. 

>It tidies and organizes the data. It uses a pivot_wider and interpolation approach to get the Actinic_par and OD values in line rowwise. This requires careful 'arrange' of the rows. It imports a metadata catalog (Data/RawData/"CultureCatalog.Rds") and merges the metadata with the imported data based
upon shared values for the variables 'MC', 'Tube', and 'Filename' which should unambiguously identify a given growth trajectory measured at OD680 or OD720. OD680 of 0.005 and OD720 of 0.003 were used to replace measured OD values for tubes with OD680 and/or OD720 < 0 across the whole growth trajectory. Assigning these very low OD values acts as a place holder; otherwise data for these tubes get filtered out later and 'no growth' data are lost [@genge_import_mcdata_2024].

>It generates preliminary data plots. It filters the data for outliers by screening out values distant from the moving average of a window in the stream; this requires careful 'arrange' of the rows so sequential rows represent sequential time steps [@genge_import_mcdata_2024].

> This works because the OD680 & OD720 data are only episodically, but widely, aberrant when a bubble interrupts the measurement, and if the Multicultivator is running properly these bubble aberration events are rare [@genge_import_mcdata_2024].

This notebook is also used for data visualizations and statistical analyses related to the microbial degradation (synonymous with loss of mass in this context) of vegetable-based bioplastics and for associated chlorophyll analyses.

# Introduction 

According to Genge et al. (2024):

>The PSI Multicultivator is used to grow 8 x 80 ml of phytoplankton culture under a common temperature regime, with individual control of bubbling, light level, light spectral quality and photoperiod for each of the 8 culture tubes.

**Research Questions**:


# Materials and Methods
## Sampling


## Preparation of Culture Nutrient

## Photobioreactor Trials

## Data Handling

Formatted display of content from .md file on GitHub site.
Upon knitr figures will be saved to 'Figs/'
```{r setup, include = FALSE}
knitr::opts_chunk$set(warning = FALSE, error = FALSE, message = FALSE)
knitr::opts_chunk$set(fig.path='Figs/')
```

### Load Libraries
```{r load libraries, echo = FALSE, warning = FALSE, message = FALSE}
library(tidyverse)
#library(lubridate)
library(broom)
library(data.table)
library(zoo)
```

This report was generated in `R version 4.3.2` [@r_core_team_r_2023], with the packages: `tidyverse`[@wickham_welcome_2019], `lubridate` [@grolemund_dates_2011], `knitr` [@xie_knitr_2023; @xie_dynamic_2015; @stodden_knitr_2014], `broom` [@robinson_broom_2024],`data.table` [@barrett_datatable_2024], and `zoo` [@zeileis_zoo_2005].

```{r set project variables}
Project <- "B2201"

# Multicultivator data files are too large to upload in a single .zip  .zipped folders can be uploaded. 

DataIn <- file.path("..", "Data", "RawData")

DataOut <- file.path("..","Data", "ProcessedData", "ImportedMCData")

# number of rows to skip upon read csv
Skip <- 20
```


```{r read locally stored metadata from rds}
# CultureCatalog <- readRDS(file = file.path("..", "Data",  "RawData","CultureCatalog.Rds"))
```

```{r MultiCulti files}

#List files in the extracted folder with a ".asc" extension
#MCFiles <- unzip(DataIn, list = TRUE)
#MCFiles <- MCFiles[grepl(".csv$", MCFiles$Name), "Name"]

#List files in data folder
list.files(DataIn)
```

```{r set TargetFile and read csv MCMIX004, warning = FALSE, message = FALSE}

TargetFile <- "20250128_B2201_MCMIX004_123_0.csv"

# This opens a compressed file without decompressing.
# TargetData <- read_csv(unz(DataIn, TargetFile),  
#                        skip = Skip, 
#                        id = "Path", 
#                        col_names = c("key", "time", 
#                                      "abs-time", "value")) %>% 
#   separate(col =Path, into = c("Path", "Filename"), sep = ":") 

TargetData <- read_csv(file = file.path(DataIn, TargetFile), skip = Skip, 
                        id = "Path", 
                        col_names = c("key", "time", 
                                      "abs-time", "value")
                       )

# %>% 
#    separate(col =Path, into = c("Path", "Filename"), sep = ":")

TargetFileName <- str_remove(string = TargetFile, pattern = ".csv")

TargetData <- TargetData |>
  mutate(Filename = TargetFileName)

```

```{r tidy MultiDataTarget MCMIX004}
#filter superfluous rows to simplify later pivot
TargetData <- TargetData %>%
  filter(str_detect(key, "od-720|od-680|actinic-lights.light")) |>
  dplyr::select(key, time, `abs-time`, value, Filename) |>
  mutate(abs_time = lubridate::dmy_hms(as.character(`abs-time`)),
         Tube = as.numeric(str_extract(key, "[:digit:]"))
         ) |>
  dplyr::select(-`abs-time`)


#extract StartHour dynamically from first row of abs_time and display for cross check
StartHour <- as.numeric(format(TargetData$abs_time[1], format = "%H"))

StartDate <- TargetData$abs_time[1]

#Generate ToD as mod 24 of time + StartHour
TargetData <- TargetData %>%
  mutate(ToD = (time + StartHour) %% 24,
         Day = round((time/24), digits = 0))
#had trouble extracting 'Day' from abs_time, only got 3 unique values
  
#extract ExpDate for matching with Catalog
# TargetData <- TargetData %>% 
#     mutate(ExpDate = str_extract(Filename, "202[:digit:][:digit:][:digit:][:digit:][:digit:]_"),
#            ExpDate = ymd(str_extract(ExpDate, "202[:digit:][:digit:][:digit:][:digit:][:digit:]")))

#extract MC for matching with Catalog
#fixed to run with MC or MCMIX.
#fixed to work for RUN #'s > 100. MC extracted from filename when Runs >99 returns the MC name and RUN #, this removes '_RUN#'
# TargetData <- TargetData %>% 
#     mutate(MC = str_extract(Filename, "MC.*[:digit:][:digit:][:digit:]"),
#            MC = str_remove(MC, pattern = "\\_.*"))
```

Data validation plot
```{r data plot}
TargetData |>
  ggplot() +
  geom_point(aes(x = abs_time, y = value)) +
  facet_grid(cols = vars(Tube))
```

# Results and Discussion

## Suprise results!? Does this plastic inhibit microbial growth?

>Plots all OD values. True detection is OD680 and false detection is OD720 [@genge_import_mcdata_2024].

```{r prelim plot MCMIX004}

 TargetData %>%
  filter(grepl("od-", key)) %>%
  ggplot(data = .) +
  geom_point(aes(x = time, y = value, colour = as.factor(str_detect(key, "680"))), size = 0.5) +
   coord_cartesian(ylim = c(-0.08,0.05)) +
  geom_vline(aes(xintercept = 4.3 * 24), 
             linetype = "dashed") +
  geom_vline(aes(xintercept = 11.3 * 24), 
             linetype = "dashed") + 
  geom_vline(aes(xintercept = 18.3 * 24), 
             linetype =  "dashed") + 
  geom_vline(aes(xintercept = 25.3 * 24), 
             linetype =  "dashed") +
  geom_vline(aes(xintercept = 32.3 * 24), 
             linetype = "dashed") +
  scale_colour_manual(values = c("black", "red")) +
  labs(y = "Optical Density (OD)", x = "Elapsed Time (h)", title = "Tubes",
       caption = "Figure 5. Optical density (OD) measurements over time at 25°C for tubes with different experimental conditions. Tubes 1–5 \nare replicates containing culture, nutrients, and plastic. Tube 6 contains culture and nutrients but no plastic. Tube 7 \ncontains nutrients and plastic but no culture. Tube 8 contains plastic but no culture or nutrients. True detections (OD680, red) \nand false detections (OD720, black) are shown. Dashed vertical lines represent when resuspension occured at \napproximately 4, 11, 18, 25, and 32 days. The y-axis is constrained to OD values between -0.1 and \n0.1 for clarity.") +
  facet_grid(cols = vars(as.factor(Tube))) +
  theme_bw() +
  theme(axis.text.x = element_text(size = 7, angle = 90), 
        plot.caption = element_text(size = 8, hjust = 0),
        plot.margin = margin(10, 10, 50, 10))


```

# Generate par_ue column with rows aligned with OD measures [@genge_import_mcdata_2024]

# Pivot_wider to get actinic-lights data aligned with relevant sensor data. Need to include arrange(Filename, time, Tube) to keep things aligned! Need to group_by and/or reorder rows appropriately; Be Careful [@genge_import_mcdata_2024]

```{r pivot_wider}
#possible issue with data structure; there are multiple values for some of the rows of actinic light columns, so the column becomes a list.
#Can add  values_fn = 
#to arbitrarily take the max or min etc. element of the list; but there might be a wider problem here when importing multiple files

TargetDataWide <- TargetData %>%
  pivot_wider(names_from = key, values_from = value, values_fn = list(value = max)) %>%
  arrange(Filename, Tube, time)

#rm(TargetData)
```

>Actinic light values do not align time wise with OD measures [@genge_import_mcdata_2024].

>Interpolate NA in actinic light columns from last observation, arrange by MC & Tube Then generate Actinic_par summary column. If multiple lights are activated, this chunk will give the summed par of all different colours for the tube. If a single actinic light is activated per tube, this gives the par for that tube. Filter rows where !is.na(Actinic_par) to check for incorrect row sums [@genge_import_mcdata_2024].

> Interpolation for Sine is not necessarily appropriate interpolation for Square photoregime; issues with propagating last Actinic_par of afternoon through evening, or back-casting first Actinic_par of morning [@genge_import_mcdata_2024].

>Small glitching adding actinic_light values for tubes where actinic_light column should be 0; issue with interpolation we think [@genge_import_mcdata_2024].

```{r interpolate and summarize actinic_par by tube MCMIX004}
#http://publish.illinois.edu/spencer-guerrero/2014/12/11/2-dealing-with-missing-data-in-r-omit-approx-or-spline-part-1/
#https://dplyr.tidyverse.org/dev/articles/colwise.html
#Interpolation causes problems with final rows that repeat last value.

interpolate <- function(x){zoo::na.locf(x, na.rm = FALSE, fromLast = FALSE, type = "l", maxgap = Inf)}

#possible problem with actinic_par for MC data b/c actinic-lights.light1 = NA generates actinic_par of 0 b/c in rowSums na.rm = TRUE, which treats NA as 0.
#possibly not a big problem but watch for bugs
#na.rm = FALSE fails to run
TargetDataWide <- TargetDataWide %>%
  group_by(Tube) %>%
  arrange(Filename, Tube, time) %>%
  mutate(across(.cols = starts_with("actinic-lights.light"), .fns = interpolate)) %>%
  ungroup() %>%
  mutate(Actinic_par = rowSums(.[grep("actinic-lights.light", names(.))], na.rm = TRUE)) %>%
  filter(!is.na(Actinic_par)) %>%
   dplyr::select(!contains("actinic-lights.light"))
```

>Now that Actinic_par is aligned with each row, coalesce od-sensors-X.od-720 and od-sensors-X.od-680 into 2 columns, b/c 'Tube'is already a column, so no need to identify tube X in od-sensors-X.od-680 columns. This might cause problems later matching OD measures to actinic light colours [@genge_import_mcdata_2024].

```{r consolidate OD}
TargetDataWide <- TargetDataWide  %>%
   mutate(OD680 = rowSums(.[grep("od-680", names(.))], na.rm = TRUE),
          OD720 = rowSums(.[grep("od-720", names(.))], na.rm = TRUE)) %>%
   dplyr::select(!contains("od-sensors"))
```

# Filter out rows where OD680 = 0 b/c time is measure more often than OD.
```{r filter OD}
TargetDataWideFilter <- TargetDataWide |>
  filter(OD680 != 0)
```



```{r metadata merge MCMIX004}
#This generates 'NA' values for ~1,000,000 rows of 'O2'; possibly the temperature rows?
# TargetDataMeta <- left_join(x = TargetDataWide, y= CultureCatalog, by = c("MC", "Tube"))
# rm(TargetDataWide)
```

No statistically significant differences in relative fluorescence units were observed (Tables 1–2). However, slight visual trends in the optical density readings (Figures 6–8, 10–12), suggests the need for further testing. Increasing the number of control replicates and extending the observation period may help determine if these trends are consistent and potentially meaningful.

```{r data validation plot}
TargetDataWideFilter %>% 
  ggplot() + 
  geom_point(aes(x = time, 
             y = OD680), size = 1, alpha = 0.5) + 
  labs( title = "680 nm Optical Density Readings of Chlorella Cultures", 
        subtitle = "90 umol photons m-2 s-1", x = "Elapsed Time (hours)", 
        y = "Optical Density (OD at 680 nm)", 
       ) + 
  facet_grid(cols = vars(Tube)) + 
  # scale_x_continuous(breaks = seq(0, 800, by = 48)) + 
  # scale_y_continuous(limits = c(0, 0.1)) +
  theme_bw() + 
  theme( axis.text.x = element_text(size = 9, angle = 90), 
         strip.text = element_text(size = 10, face = "bold"),
         plot.caption = element_text(size = 8, hjust = 0),
         plot.margin = margin(10, 10, 50, 10) )
```



## Microbes Present 

## Limitations

## Future Direction


# Conclusion


# References


